Differential Expression

### 2020-04-07 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
DE contrast for HBCC DLPFC

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
                                 parentId ="syn21887806",
                                 folderName = "Differential Expression - Dx.Tissue")
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21887806';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Dx.Tissue';

thisFileName <- 'Differential-Expression.Rmd'

@@

# Download CQN counts
CQN_ID = "syn21888972"
ALL_USED_IDs = c(CQN_ID)
CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id") %>% 
  data.matrix()

# Download filtered counts
COUNTS_ID = "syn21888969"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
NEW.COUNTS = downloadFile(COUNTS_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Get covariates
COVAR_ID = "syn21888968"
ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
COVARIATES = downloadFile(COVAR_ID) %>% 
  tibble::column_to_rownames(var = "SampleID")

# Set variable class
FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue", "IsolationBatch")
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx.Tissue", "RIN", "IntragenicRate")
```

### Normalisation - Dx.Tissue
1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect

@@

### Tissue Correlation
```{r tissue_correlation}
hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  dplyr::select(Individual_ID) %>% 
  table

df_stg = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("STG", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("ACC", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)

df_merge = merge(df_stg, df_acc, by="Individual_ID")

res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")

corRes = sapply(res_std$ensembl_gene_id, function(ensGene){

  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]

  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
})

hist(corRes, xlab="Correlation", main="Correlation between ACC and STG sample from the sample individual for each gene")

pca = prcomp(t(VOOM.GENE_EXPRESSION$E))

df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), 
               by.x="row.names", 
               by.y="SampleID" ) %>% 
  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_") %>% 
  dplyr::mutate(Institution = "NIMH-HBCC")

ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
  dplyr::mutate(Institution = "NIMH-HBCC")

@@

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'), drop = F])

@@

```{r diffExp, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))

L = cbind(L1, L2, L3, L4, L5, L6)

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"),
              glue::glue_collapse(names(L4[L4 != 0]), "-"),
              glue::glue_collapse(names(L5[L5 != 0]), "-"),
              glue::glue_collapse(names(L6[L6 != 0]), "-")
)

@@

# Get gene lengths
counts = 'syn21867866'
ALL_USED_IDs = counts
counts = downloadFile(counts) %>% data.frame()
gene_lengths = counts[,c("Geneid", "Length")] %>% 
  dplyr::rename(gene_id = Geneid)
rm(counts)

@@

# Get differential expression
DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

  tissue	= 'superior temporal gyrus, anterior cingulate cortex', 
  study = 'CMC',

@@

# Code
CODE <- Folder(name = "Differential Expression - Dx.Tissue", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', 
                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Dx.Tissue_dream-output.rds')
ADJ_OBJ = File('CMC_ACC_STG_Dx.Tissue_dream-output.rds',
               name = 'Dream Output object - Dx.Tissue', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```

>>>>

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
                                 parentId ="syn21859863",
                                 folderName = "Differential Expression - Dx")
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21859863';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Dx';

thisFileName <- 'Differential-Expression_Dx.Rmd'

@@

# Download CQN counts
CQN_ID = "syn21862457"
ALL_USED_IDs = c(CQN_ID)
CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id") %>% 
  data.matrix()

# Download filtered counts
COUNTS_ID = "syn21862455"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
NEW.COUNTS = downloadFile(COUNTS_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Get covariates
COVAR_ID = "syn21862454"
ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
COVARIATES = downloadFile(COVAR_ID) %>% 
  tibble::column_to_rownames(var = "SampleID")

# Set variable class
FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx", "IsolationBatch")
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx", "RIN", "rRNARate", "IntergenicRate")
```

### Normalisation - Dx
1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect

@@

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(Tissue = "DLPFC") %>%  
  dplyr::mutate(Institution = "NIMH-HBCC")

@@

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx'), drop = F])

@@

```{r diffExp, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("DxOther", "DxControl"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("DxSCZ", "DxControl"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("DxSCZ", "DxOther"))

L = cbind(L1, L2, L3)

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-")
)

@@

# Get gene lengths
counts = 'syn21754352'
ALL_USED_IDs = c(ALL_USED_IDs, counts)
counts = downloadFile(counts) %>% data.frame()
gene_lengths = counts[,c("Geneid", "Length")] %>% 
  dplyr::rename(gene_id = Geneid)
rm(counts)

@@

# Get differential expression
DE = lapply(glue::glue("L{1:3}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

  tissue	= 'dorsolateral prefrontal cortex', 
  study = 'CMC',

@@

# Code
CODE <- Folder(name = "Differential Expression - Dx", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_Dx_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_HBCC_DLPFC_Dx_netResidualExpression.tsv', 
                name = 'Dx Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_HBCC_DLPFC_Dx_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_HBCC_DLPFC_Dx_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Dx
saveRDS(ADJUSTED.FIT, file = 'CMC_HBCC_DLPFC_Dx_dream-output.rds')
ADJ_OBJ = File('CMC_HBCC_DLPFC_Dx_dream-output.rds',
               name = 'Dream Output object - Dx', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```
